---
title: "GR5065 Homework 1"
date: "Due February 7, 2022 at 6PM"
author: "Ben Goodrich"
editor_options: 
  chunk_output_type: console
urlcolor: blue
output: 
  pdf_document: 
    latex_engine: xelatex
    number_sections: yes
header-includes:
   - \usepackage{amsmath}
   - \usepackage{color}
   - \usepackage{pst-poker}
   - \usepackage{cancel}
---

# Poker

## Probability of rare five-card combinations

Before the hand starts (i.e. irrespective of any betting and presuming
the player does not fold)

- What is the probability of a player ending up with a "flush", which is defined 
as a five-card combination where all five cards are of the same suit? You can
ignore the tiny probability of obtaining a "straight flush", where all five cards
have adjacent values in addition to being of the same suit.
- What is the probability of a player ending up with a "full house" --- which is
defined as a five-card combination where three of the cards are of the same
value and the other two cards have the same value --- but not four-of-a-kind?

For the flush,
```{r}
( dhyper(x = 5, m = 13, n = 52 - 13, k = 7) +
  dhyper(x = 6, m = 13, n = 52 - 13, k = 7) + 
  dhyper(x = 7, m = 13, n = 52 - 13, k = 7) ) * 4
```
The first line calculates the probability of exactly five out of seven cards
are of the same suit. But it is also a flush in the unlikely event that all
six or seven cards of the same suit (only five of which count), so we use the 
General Addition Rule with disjoint events on the next two lines. Finally, we 
multiply by four because there are four suits that each have the same probability 
of comprising a flush (and ignore a few strait flushes).

For the full house, there are three ways the seven cards could be allocated:

(A) 3 of a one type of card, 3 of a second type of card (one of which does not count 
toward the five), and 1 of a third type of card (otherwise it would be four-of-a-kind).
Thus, there is a ${13 \choose 2}$ factor below, because there are $13$ values of
cards that we are selecting two of to comprise the full house, and a factor of $44$
because there are $44$ cards of the third type.
(B) 3 of one type of card, 2 of a second type of card, and 2 of a third type
of card (neither of which counts toward the five). Thus, there is a factor of $13$
below because there are $13$ values of cards to obtain three of and then a 
${12 \choose 2}$ factor because there $12$ other values of cards that we need
two pairs of.
(C) 3 of one type of card, 2 of a second type of card, and 2 of distinct types of cards
(neither of which counts toward the five). By the same reasoning as previously,
there is a factor of $13$ below, a factor of $12$, and a factor of ${11 \choose 2}$

Thus, we have
```{r}
total <- choose(52, 7)
A <- choose(4, 3)^2 * 44 * choose(13, 2) / total
B <- choose(4, 3) * choose(4, 2)^2  * 13 * choose(12, 2) / total
C <- choose(4, 3) * choose(4, 2) * 4^2   * 13 * 12 * choose(11, 2) / total
A + B + C
```

The probability of a flush or a full house and other hands are listed at

https://en.wikipedia.org/wiki/Poker_probability#Frequency_of_7-card_poker_hands

which also provides some additional detail on the combinatorics of how they are 
calculated using the choose function.

## Pre-flop action

Before the hand starts, what number is the probability of Selbst being dealt two Aces 
as hole cards? Then, for each of the following explain your thinking (but exact
calculations are not required):

- From Selbst's perspective when she made the decision to bet the first time, is the 
probability that she has two Aces higher or lower than the probability of being dealt two Aces?
- From Baumann's perspective when she made the decision to call the first time, is the 
probability that Selbst has two Aces higher or lower than the probability of being dealt two Aces?
- From Rosen's perspective when he made the decision to fold, is the probability that 
Selbst has two Aces higher or lower than the probability of being dealt two Aces?

The probability of being dealt two Aces is
```{r}
4 / 52 * 3 / 51 # = dhyper(x = 2, m = 4, n = 52 - 4, k = 2)
```
or $\frac{1}{13} \times \frac{1}{17} = \frac{1}{221}$.

From Selbst's perspective the "probability" that she has two Aces is $1$, which is
to say that she knows it with certainty because she sees the \As \Ad \ with her eyes.
This seems trivial but it is actually an important point in Bayesian analysis: A
realization of a random process is data to condition on if you observe it.

Baumann cannot see that Selbst has two Aces but does observe that Selbst bets, which
is consistent with Selbst having two Aces but is also consistent with Selbst having
about $\frac{1}{7}{52 \choose 2}$ other good hands. By Bayes Rule, the conditional 
probability that Selbst has a pair of Aces given that Selbst bets is

$$\Pr\left(\mbox{Aces} \mid \mbox{bets}\right) =
\frac{\frac{1}{221} \times 1}{\frac{1}{7}} = \frac{7}{221} \approx 0.032,$$

which reads as the prior probability from above that Selbst is dealt a pair of Aces,
times the conditional "probability" that Selbst bets given that she has
a pair of Aces (which is guaranteed), divided by the probability that
Selbst bets irrespective of whether she has a pair of Aces, which was
said to be roughly $\frac{1}{7}$. Technically, the calculation should
condition on the fact that Baumann has two sevens, so there are four Aces
among the $50$ cards left in the deck, and Selbst would have bet with 
a few hands involving Baumann's \sevh \sevd, but those considerations would not 
change the posterior probability very much, and it would still be considerably 
greater than the prior probability of $\frac{1}{221}$.

From Rosen's perspective, the fact that he has the Ace of hearts is
much more consequential to the calculation. Now Bayes Rule implies
$$\Pr\left(\mbox{Aces} \mid \mbox{bets}, \Ah\right) =
\frac{\frac{3}{51} \times \frac{2}{50} \times 1}{\frac{1}{7}} \approx 0.016,$$
where we have recomputed the prior probability that Selbst is dealt
two Aces from a 51-card deck that is missing the \Ah \ (but ignoring
the fact that it is also missing the \sixc) but again multiplied
by the conditional "probability" that Selbst would bet with a pair
of Aces and divided by the probability that Selbst would bet irrespective
of whether she has a pair of Aces (and ignoring that Selbst would have
bet with several hands involving Rosen's \Ah). This posterior probability is
still higher than the probability of Selbst being dealt a pair of
Aces but is considerably lower than the probability from Baumann's perspective.

## Schwartz's call

Explain in words why it is a good decision (exact calculations are
not required on this subproblem) for Schwartz to call pre-flop.

By betting $250$ more chips, Schwartz gives himself a chance to win the
pot which at that moment is $1,025$ chips. The expected change in 
chips would be

$$1025 \times \Pr\left(\mbox{wins} \mid \dots\right) - 250 \times 
\left(1 - \Pr\left(\mbox{wins} \mid \dots\right)\right)$$

So, Schwartz only needs about a $\frac{1}{5}$ chance of 
winning the hand in order for this expected change to be positive
and thus exceed the value of zero chips gained if Schwartz were
to fold pre-flop. Even though both Selbst and Baumann both presumably
have very good hands, almost any two decent hole cards would give Schwartz
at least a $\frac{1}{5}$ chance of winning the hand. It is more complicated
because there will be more betting post-flop, but presuming Schwartz
will make optimal decisions post-flop, it also makes sense for him to call
pre-flop.

## Flop action

If Schwartz's utility function were equal to his chips
at the conclusion of the hand and Schwartz would be indifferent between folding and 
calling Selbst's bet of $700$ chips (which Baumann called, bringing the pot to $2,675$ chips) 
if he had the \Qc \ and \eigh, what must the conditional probability be that either Selbst or 
Baumann has the \Kc? To simplify this subproblem, you can assume that Schwartz would 
definitely win the pot if he makes a club flush and no one else has the \Kc \ (although that 
turned out not to be true in this video) and that Schwartz would certainly lose if he does not 
make a flush. Also, you can assume that there would be no more betting.

Since Schwartz can see his two hole cards and three clubs in the middle, if
either of the next two cards is the \Kc, then Schwartz would know he has the 
best possible flush with the \Qc. The probability of that is
```{r}
dhyper(x = 1, m = 1, n = 52 - 2 - 3, k = 2)
```
Alternatively, if either (or both) of the next two cards is a club but not 
the \Kc, then Schwartz would know he has the second-best possible flush. The
probability of that is
```{r}
dhyper(x = 1, m = 13 - 5, n = 52 - 8 - 1, k = 2) +
dhyper(x = 2, m = 13 - 5, n = 52 - 8 - 1, k = 2)
```
If you add those three terms together, ignore the chance that either Selbst
or Baumann ends up with a hand that is better than a flush, and round, you obtain 
the 33% chance of Schwartz winning that you see on the video at 0:57.

But Schwartz does not know that neither Selbst nor Baumann has the \Kc, so the
last two terms need to be multiplied by the conditional probability, say $C$, that 
neither Selbst nor Baumann has the \Kc \ when calculating the expected change 
in chips as

$$2675 \times 0.0417 + 2675 \times 0.2917 \times C - 
700 \times (1 - 0.0417 - 0.2917 \times C)$$
Set that equal to zero because Schwartz is supposedly indifferent between
calling and folding and solve for $C \approx 0.568$. Thus, the conditional
probability that either Selbst or Baumann has the \Kc \ is about $0.432$ or
perhaps a little lower if we were to take into account that either Selbst or 
Baumann could end up with a hand that is better than a flush. Intuitively,
Selbst or Baumann would be betting after the flop if either had the \Kc, but they 
would also be betting with a somewhat larger collection of hands that do not
involve the \Kc.

## Pre-turn calculation

After Schwartz folds at 1:02, the percentage chances (in the bottom left of the video)
switch to 91% for Selbst to win and 4% for Baumann to win, with an implicit 5% chance that 
they tie if the remaining two cards in the middle are both clubs (all percentages were rounded). 
In the event that both Selbst and Baumann end up with three-of-a-kind, Selbst would win because 
Aces are better than sevens. How did the video arrive at these percentages?

The probability that Baumann wins is the probability that either of the next
two cards in the middle is the \sevs \ and the other is not the \Ah, which is about 4%
```{r}
dhyper(x = 1, m = 1, n = 52 - 7, k = 2) * 
(1 - dhyper(x = 1, m = 1, n = 52 - 8, k = 1))
```
The probability that both of the next two cards are clubs is about 5% when
there are $10$ clubs left in a deck of $45$ cards
```{r}
10 / 45 * 9 / 44 # = dhyper(x = 2, m = 10, n = 35, k = 2)
```
And if anything else happens, then Selbst would win which is the remaining 91%.

## Turn

The next card revealed in the middle (called the "turn") is \sevs \ and the commentators 
go crazy because Selbst now has a full house but Baumann has four-of-a-kind. Before
the hand started, irrespective of any betting and presuming neither player folds, what was
the probability of one player at the table ending up with four-of-a-kind and another
player ending up with a full house?

We can break down the probability of someone getting four of one type of card
```{r}
dhyper(x = 4, m = 4, n = 52 - 4, k = 7)
```
as

(A) The probability that the player has a pair and the board has that same pair
(B) The probability that the board has three of some type card and the player has the other one
(C) The probability that the board has four of that type of card
```{r}
A <- dhyper(x = 2, m = 4, n = 48, k = 2) * dhyper(x = 2, m = 2, n = 48, k = 5)
B <- dhyper(x = 1, m = 4, n = 48, k = 2) * dhyper(x = 3, m = 3, n = 47, k = 5)
C <- dhyper(x = 0, m = 4, n = 48, k = 2) * dhyper(x = 4, m = 4, n = 46, k = 5)
A + B + C
```

In scenario (C), it is impossible for another player to have a full house because both
players have four-of-a-kind. In scenario (A), the other player would need to have three
of a kind among their two hole cards and the three other cards on the board. In
scenario (B), the other player would need to have a pair among their two hole cards
and the two other cards on the board. Thus, the joint probability is

```{r}
D <- A * dhyper(x = 3, m = 4, n = 44, k = 5) +
     B * dhyper(x = 2, m = 4, n = 44, k = 4)
```

Furthermore, `D` must be multiplied by $13$, $12$, and $2$ because there are
$13$ ways for the first player to make four-of-a-kind, $12$ ways each for the 
second player to make a full house, and $2$ orderings of a pair of players. 
The probability of one player making a four-of-a-kind and the other making
a full house is

```{r}
D * 13 * 12 * 2
```

which is perhaps greater than one might think, but much of that involves
the less dramatic scenario (B) rather than the scenario (A) that happens in the video.

Taking into account that there are eight players at the table would be more 
difficult, but usually all but two players fold pre-flop anyway.

It is possible to check your answer with a simulation, but you need to
draw many times in order to estimate small probabilities with high
relative accuracy.
```{r, sim, cache = TRUE, message = FALSE}
library(dplyr)
library(parallel)
deck <- rep(1:13, each = 4)
S <- 10^6
mclapply(1:S, mc.cores = detectCores(), FUN = function(s) {
  cards <- sample(deck, size = 2 * 2 + 5, replace = FALSE)
  board <- tail(cards, n = 5)
  if (n_distinct(board) == 5) return(FALSE) # full house is impossible
  hole_cards <- matrix(head(cards, n = -5), nrow = 2)
  four <- apply(hole_cards, MARGIN = 2, FUN = function(hand) {
    any(table(c(hand, board)) == 4)
  })
  if (sum(four) != 1) return(FALSE) # no isolated four-of-a-kind
  full_house <- apply(hole_cards, MARGIN = 2, FUN = function(hand) {
    tab <- table(c(hand, board))
    all(2:3 %in% tab) | sum(tab == 3) == 2
  })
  if (!any(full_house)) return(FALSE) # no full house
  # Are the four-of-a-kind and the full house not the same player?
  !identical(which(four), which(full_house))
}) %>% unlist %>% mean
```

## River action

Explain with reference to Bayes Rule why Selbst's decision to call Baumann's raise was 
justifiable. To do so, calculate the conditional probability that Selbst wins the
pot given the seven cards that Selbst can see and all the previous betting. Note that
Selbst's decision to call can be justified even if the probabilities that you use are not
exactly correct.

Conditioning on the fact that Baumann raised all-in eliminates a lot of hands that she
could have from Selbst's consideration, such as a flush. In other words, 
$\Pr\left(\mbox{All in} \mid \mbox{flush}\right) = 0$ because from Baumann's perspective
it would be too likely that Selbst has a hand that is better than a flush and
too unlikely that Selbst would call with a hand that is worse than a flush. Although
Baumann might have raised all-in with \Ah \sevd, conditioning on the fact that Baumann
called pre-flop makes it very unlikely that Baumann has an Ace and a seven of different
suits.

If Baumann would definitely raise all-in with \Ah \sevh, then the conditional probability
that Baumann has \Ah \sevh \ would be the same as the conditional probability that
Baumann has \sevh \sevd. In other words, both hands are equally consistent with
the available information. Bayes Rule implies

$$\Pr\left(\mbox{Selbst wins} \mid \dots\right) =
\frac{\Pr\left(\mbox{Baumann has \Ah \sevh \ and raises all-in}\right)}
{\Pr\left(\mbox{Baumann has \Ah \sevh \ and raises all-in}\right) + 
 \Pr\left(\mbox{Baumann has \sevh \sevd \ and raises all-in}\right)} = \frac{1}{2}$$
If there were any other hole cards that Baumann would raise all-in with, the
probability that Selbst would win would be even higher than $\frac{1}{2}$.

Thus, if Selbst has at least a $\frac{1}{2}$ chance to win $66,975$ chips
by risking losing another $20,300$ chips, the expected change in chips
if Selbst's calls is very positive. In order for the expected change in 
chips to be negative, Selbst would have to believe that the probability
that Baumann would raise all-in with \Ah \sevh \ is much lower than
the "probability" that Baumann would raise all-in with \sevh \sevd,
which is a certainty. It seems from the video that Selbst is not entirely sure whether
Baumann would raise all-in with \Ah \sevh \ but that probability would have 
to be lower than $\frac{1}{3}$ to make calling a bad decision.

## Baumann's strategy on the river

Given that the five cards in the middle are \Ac \sevc \fivec \sevs \fourd, show that if Baumann's 
strategy were to raise all-in only if she either had two sevens or two Aces as hole cards,
_and if Selbst knows that_ then Baumann's strategy would have a worse expectation than
a strategy of calling with either two sevens or two Aces. This implies that Baumann would 
also have to raise all-in with a third hand that is even worse (a bluff) in order for the 
all-in strategy as a whole to have a positive expectation.

If Selbst knows that this is Baumann's strategy, given that Baumann raises all-in, then
Baumann must have either two sevens or two Aces. Thus, if Selbst has two sevens, then Baumann
must have two Aces, in which case Selbst would win the pot by calling. Otherwise, Selbst would
fold and not lose any more chips. In short, if Selbst knows that this is Baumann's strategy,
Baumann will never get called by a worse hand and would only be called by a better hand. 
Thus, Baumann would be be better off calling in both cases rather than raising all-in in both.

# Surnames

## Notation

Rewrite the denominator in equation (2) from the paper using the crossed-out notation
like that we used for bowling, $\Pr\left(\bcancel{x_1} \bigcap x_2 \mid n = 10\right)$.

$$\Pr\left(R_i = r \mid S_i = s, G_i = g\right) =
\frac{\Pr\left(G_i = g \mid R_i = r\right) \Pr\left(R_i = r \mid S_i = s\right)}
{\sum_{r^\prime \in \mathcal{R}} \Pr\left(G_i = g \mid R_i = r^\prime\right) 
\Pr\left(R_i = r^\prime \mid S_i = s\right)} =
\frac{\Pr\left(G_i = g \bigcap R_i = r \mid S_i = s\right)}
{\Pr\left(G_i = g \bigcap \bcancel{R_i = r} \mid S_i = s\right)}$$

## Frequentist Perspective

Would Fisher approve or disapprove of this use of Bayes' Rule in the paper? Why?

On one hand, there are no theoretically unobservable parameters in this model, and
this paper does not actually use any Bayesian _estimation_ methods that we will learn
later in this semester. However, Fisher would still find this use of Bayes Rule to be
objectionable. What race person $i$ is analogous to the question of whether a huge odd 
integer is prime? There is no repeatable experiment in which person $i$ is white some 
proportion of the time and not white other times.

One could ask the question what is the probability of randomly _drawing_ a white
person from a population, which Frequentist probability could be applied to,
but that is not what is going on here. The people who register or turn out to vote in an
election are not a random sample from a well-defined population. This is basically
a missing data problem because the race of the voter is not recorded in most states.

## You

```{r, message = FALSE}
NYS <- readRDS("NYS.rds")
my_df <- data.frame(surname = c("Goodrich", # Ben
                                "Chan",     # Monica
                                "Perez"),   # Abel
                    state = "NY", county = "061", 
                    age = c(42, 27, 26), sex = c(0, 1, 0))
library(wru)
me <- predict_race(my_df, census.geo = "county", census.data = NYS, age = TRUE, sex = TRUE)
me
```

The posterior probability than Ben is white is about $0.87$, which is the
much greater than the probability that he is of any other race. Similarly
for Monica and Abel, their last names result in posterior probabilities
that are even closer to $1$, despite the fact that there are much fewer
Asians and Hispanics in Manhattan than whites.

## Granularity

Researchers can use this method with Census geolocation data at the county-level
or at smaller geographical units, such as the Census tract (which contains
about $4,000$ contiguous people). How would you anticipate the
probabilities in `me` changing if we were to use the Census tract data
(and the tract that you live in now) rather than the county-level data? Why?

Ben lives in West Harlem, which has a smaller percentage of white people
and a larger percentage of black people than New York county (i.e. Manhattan)
as a whole (although West Harlem has become more white in recent years, 
this trend would not have been picked up as much in the $2010$ Census). Also, 
knowing that Ben's last name is Goodrich does not tell as much about whether 
he is white or black as it does tell you that he is neither Hispanic, Asian, nor "other". 
Thus, if the analysis were conducted at the Census tract level, a little bit of the 
posterior probability that Ben is white would shift to black.
